---
title: "intro"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{intro}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(scPloidy)
```

## Analyzing your own data

### Using fragments.tsv.gz generated by 10x Cell Ranger

In the Cell Ranger output directory, you should have files
`fragments.tsv.gz` and `fragments.tsv.gz.tbi`.
The file `fragments.tsv.gz` can be processed with the `fragmentoverlapcount` function,
specifying the option `Tn5offset = c(1, 0)`

### Using BAM file

Although this requires several steps, you can start from a BAM file
and generate fragments file from scPloidy.
You need to install `samtools`, `bgzip` and `tabix`,
and run the following in your shell.

First generate a BAM file in which the cell barcode is prepended to read name,
separated by ':'.
For example, suppose in your input file `bap.bam`
your barcode `BCxxxxxx` was stored in the field with `DB` tag as
`DB:Z:atac_possorted_bam_BCxxxxxx`.
We generate a BAM file `snap.bam`.

```{bash, eval = FALSE}
# extract the header file
samtools view bap.bam -H > header.sam

# create a bam file with the barcode embedded into the read name
cat <( cat header.sam ) \
<( samtools view bap.bam | awk '{for (i=12; i<=NF; ++i) { if ($i ~ "^DB:Z:atac_possorted_bam_BC"){ td[substr($i,1,2)] = substr($i,25,length($i)-24); } }; printf "%s:%s\n", td["DB"], $0 }' ) \
| samtools view -bS - > snap.bam
```

We next sort the reads by barcode,
and obtain the file `snap.nsrt.bam`.

```{bash, eval = FALSE}
samtools sort -n -@ 20 -m 10G snap.bam -o snap.nsrt.bam
```

Finally, we extract fragments from the name-sorted BAM file,
and obtain the file `fragments.txt.gz`.

```{bash, eval = FALSE}
samtools view -f 0x2 -F 0x900 -q 30 snap.nsrt.bam  \
  | samtofragmentbed.pl  \
  | perl -ne 'chomp; @a=split; $a[3]=~s/:.*//; print join("\t",@a), "\n"'  \
  | LC_ALL=C sort -T ./ -S 50% --parallel=12 -k1,1 -k2,3n -k4,4 -t$'\t' | uniq  \
  | bgzip > fragments.txt.gz
tabix -b 2 -e 3 fragments.txt.gz
```

The Perl script samtofragmentbed.pl is included in this package
as this file:

```{r}
system.file("perl", "samtofragmentbed.pl", package = "scPloidy")
```

The file `fragments.txt.gz` can be processed with the `fragmentoverlapcount` function,
specifying the option `Tn5offset = c(4, -5)`
